---
title: 'STATS 191: Homework Assignment 3'
author: "Dr. Pratheepa Jeganathan"
date: "10/11/2019"
output: 
  pdf_document:
    includes:
      in_header: header.tex
bibliography: AppliedStat.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**You may discuss homework problems with other students, but you have to prepare the written assignments yourself.**

**Please combine all your answers, the computer code and the figures into one PDF file, and submit a copy to gradescope.**

\rc Please use **newpage** to write solution for each part of a question.\bc

\rc Please specify the page number for each part of a question in gradescope. \bc

<!-- \rc You cannot submit multiple answers for a question. \bc -->

**Grading scheme: $\left\lbrace 0, 1, 2\right\rbrace$ points per question, total of 40.**


\rc Due date: 11:59 PM October 18, 2019 (Friday evening).\bc


# Question 1

This question is from our textbook **CH** Exercises 2.1, Page 53.

In order to investigate the feasibility of starting a Sunday edition for a large metropolitan newspaper, information was obtained from a sample of 34 newspapers concerning their daily and Sunday circulations (in thousands) *(Source: Gale Directory of Publications, 1994)*. The data can be read from the book's Website: [\blc http://www1.aucegypt.edu/faculty/hadi/RABE5/Data5/P054.txt\bc](http://www1.aucegypt.edu/faculty/hadi/RABE5/Data5/P054.txt).

1. Read the data using `read.table` (separator of column is `tab` and the data frame has variable names).

2. Construct a scatter plot of Sunday circulation versus daily circulation. 

3. Does the plot suggest a linear relationship between daily and Sunday circulation?

4. Fit a regression line predicting Sunday circulation from daily circulation (Use `lm()`).

5. Is there a significant relationship between Sunday circulation and daily circulation? Justify your answer by a statistical test (Use F test in `anova()`). 

6. Indicate what hypothesis your are testing and your conclusion for the test in part (5).

7. Using the `anova` table produced in part (5), compute the proportion of the variability in Sunday circulation is accounted for by daily circulation. 

# Question 2 

Let $Y$ and $X$ denote variables in a simple linear regression of median home prices versus median income in state in the US. Suppose that the model
$$
Y = \beta_0 + \beta_1 X + \epsilon
$$
satisfies the usual regression assumptions.

The table below is a table similar to the output of `anova` when passed a simple linear regression model.


    Response: Y
              Df Sum Sq Mean Sq F value    Pr(>F)    
    X          1     NA    5291      NA        NA
    Residuals 48 181289      NA

1. Compute the missing values in the above table.

2. Test the null hypothesis $H_0 : \beta_1 = 0$ at level $\alpha = 0.05$ using the above table.

3. Can you test the hypothesis $H_0 : \beta_1 < 0$ using the above table? (You may need to use the relationship between the T statistic and the F statistic.)

3. What proportion of the variability in $Y$ is accounted for by $X$? 

4. If $Y$ and $X$ were reversed in the above regression, what would you expect $R^2$ to be?

# Question 3

In this problem, we will investigate what happens when the assumptions of the simple linear regression model do not hold. When generating data below, set $X$ to be equally spaced between 0 and 1 (i.e. `X = seq(0, 1, by=0.01)`) and use the regression function
$$
Y = 1 + 2 \cdot X + \epsilon
$$

1. Write a **function** (call the function as `generateTstat`) to generate data from the simple linear regression model with regression function as above and normally distributed errors $\epsilon \sim \text{N}\left(0, \sigma^{2}\right)$ (can use $\sigma^{2} = 1$), returning the $T$-statistic for testing whether the slop of the regression line is equal to 2. 
[That is, testing $\text{H}_{0}: \beta_{1} = 2$ versus $\text{H}_{a}: \beta_{1} \neq 2$.]

The function arguments should be values for $X$ and slope of the regression line `beta1`.

```{r}
generateTstat = function(X, beta1){
  #Y = write Y as a function of X and error 
  #fit = fit regression line using lm 
  # beta1hat = compute least squares estimate of slope using summary(fit)$coefficient[2,1]   
  # se_beta1hat = compute standard error of slope estimate using summary(fit)$coefficients[2, 2]
  # Tstat = Compute the T-statistic using the appropriate formula (for testing H0: beta_1 = 2)
  # return(Tstat)
}
```

  (I) Using your function, run a simulation with 5000 repetitions to see if the $T$-statistic has distribution close to a $T$ distribution. How many degrees of freedom should it have (consider the length of $X$ to answer this question)? 

```{r}
# X = use seq() to generate 100 X between 0 and 1
# t_stat_vec = use replicate() to compute 5000 T-statistic values
```

```{r}
#Plot the distribution of the T-statistic and the T distribution
```

  (II) In part (I), how often is your $T$ statistic larger than the usual 5\% threshold?
```{r}
#threshold = find the threshold for testing H0: beta_1 = 2 versus Ha: beta_1 not equal to 2 
#(degrees of freedom depends on the length of X)

# Find how many of absolute t_stat_vec is greater than the threshold
```


2. Write a new function with the same regression function but errors that are t-distributed using, say, `rt` with 5 degrees of freedom to generate errors. Repeat (I) and (II) in part (1). Does the $T$-statistic still have close to a $T$ distribution? How often is your $T$ statistic larger than the usual 5% threshold? 


3. Write a new function with same regression function but errors that do not have the same variance though they are normally distributed. Construct errors such that the variance of the $i$-th error is `1+X[i]` (recall that `X` is equally spaced over interval 0 to 1 and you can specify in `rnorm` what is the standard deviation of error). Plot the variance of error as a function of `X`. Repeat (I) and (II) in part (1). Does the $T$-statistic still have close to a $T$ distribution? How often are your $T$ statistics larger than the usual 5% threshold?


4. Write a new function with same regression function but errors that do not have the same variance though they are normally distributed. Exaggerate the effect of non-constant variance by making the variance of errors `exp(1 + 5 * X[i])`. Plot the variance as a function of `X`. Repeat (I) and (II) in part (1). Does the $T$-statistic still have close to a $T$ distribution? How often are your $T$ statistics larger than the usual 5% threshold?

5. Write a new function with same regression function but errors that are not independent. Do this by first generating a vector of errors `error` and then returning a new vector whose first entry is `error[1]` but for $i>1$ the $i$-th entry is `error[i-1] + error[i]`. Repeat (I) and (II) in part (1). Does the $T$-statistic still have close to a $T$ distribution? How often is your $T$ statistic larger than the usual 5% threshold?

6. Summarize your findings in questions 1-5. Which of the departures from the assumptions for the error term of the simple linear regression model seem important?


