---
title: "STATS 191: Introduction to Applied Statistics"
author: ""
date: ""
output: 
  pdf_document:
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  Instructor 
Pratheepa Jeganathan

- E-mail: pjeganat [at] stanford [dot] edu; **Write in the subject line STATS 191**.
- Office: 390 Serra Mall, Sequoia Hall Room 116.
- Office hours: Wednesday 3.00 p.m. - 4.30 p.m. in 105 Sequoia or by an email appointment, or Monday 11.00 a.m. - 12.30 p.m. Zoom meeting.
- Course website: [Canvas @ Stanford University ](https://canvas.stanford.edu/courses/108253), this will contain a bulletin board, lecture slides, homework assignments, reading list, R, R Studio, R Markdown resources etc.

##  Teaching Assistants & Office Hours
<!-- Qian Zhao -->

<!-- - Office hours: Thursdays 8 a.m. - 10 a.m., Bowker 207. -->

<!-- Chenyang Zhong -->

<!-- - Office hours: Fridays 8 a.m. - 10 a.m., location TBD -->


##  Course Information
- Term: Autumn 2019
- Units: 3
- Time: Mon, Wed, Fri 1:30 PM - 2:20 PM
- Location: [Gates B3](https://campus-map.stanford.edu/?srch=Gates+B3#) 
- LEC: 09/23/2019 - 12/06/2019 (10 Weeks - 30 hours)

##  Course Overview

Statistical tools for modern data analysis. Topics include regression and prediction, elements of the analysis of variance, bootstrap, and cross-validation. Emphasis is on conceptual rather than theoretical understanding. Applications to social/biological sciences. Student assignments require use of the software package [R](http://cran.r-project.org). 

## Expected outcomes

By the end of the course, students should be able to: 

- Enter tabular data using [R](http://cran.r-project.org).
- Plot data using [R](http://cran.r-project.org), to help in exploratory data analysis.
- Formulate regression models for the data, while understanding some of the limitations and assumptions implicit in using these models.
- Fit models using [R](http://cran.r-project.org) and interpret the output.
- Test for associations in a given model.
- Use diagnostic plots and tests to assess the adequacy of a particular model.
- Find confidence intervals for the effects of different explanatory variables in the model.
- Use some basic model selection procedures, as found in [R](http://cran.r-project.org), to find a *best* model in a class of models.
- Fit simple ANOVA models in [R](http://cran.r-project.org), treating them as special cases of multiple regression models.
- Fit simple logistic and Poisson regression models.

## Prerequisites
An introductory statistics course, such as
- [STATS 60](https://explorecourses.stanford.edu/search?view=catalog&filter-coursestatus-Active=on&page=0&catalog=&academicYear=20182019&q=STATS+60%3A+Introduction+to+Statistical+Methods%3A+Precalculus+&collapse=) or [STATS 110](https://explorecourses.stanford.edu/search?view=catalog&filter-coursestatus-Active=on&page=0&catalog=&academicYear=20182019&q=STATS+110%3A+Statistical+Methods+in+Engineering+and+the+Physical+Sciences&collapse=) or [STATS 141](https://explorecourses.stanford.edu/search?view=catalog&filter-coursestatus-Active=on&page=0&catalog=&academicYear=&q=stats141&collapse=), and [R](https://cran.r-project.org/).

##  Texts
- Required:
  - **(CH)** [Regression Analysis by Example](http://www1.aucegypt.edu/faculty/hadi/RABE5).
    - Authors: Samprit Chatterjee, Ali S. Hadi
    - Edition: $5^{th}$ Edition
    - Print ISBN:978-0-470-90584-05 
    
##  Software
- In this course, we will use R for computing and R Markdown for producing lecture slides, solutions for homework assignments. R Markdown is highly recommended to write the solutions for homework assignments. Install the following software:
    - R (required): [https://www.r-project.org/](https://www.r-project.org/).
    - R Studio is highly recommended for syntax highlighting, package management, document generation, and more: [https://www.rstudio.com/](https://www.rstudio.com/).
        - The newest version of R Studio is highly recommended (v1.1.463).
    - Latex, which will enable you to create PDFs directly from the R Markdown in RStudio.
        - Mac users should download macTeX [http://www.tug.org/mactex/downloading.html](http://www.tug.org/mactex/downloading.html) from Safari (not Chrome).
        - Windows users should install MiKTeX [https://miktex.org/download](https://miktex.org/download).

##  Grading
The final letter grade for this course will be determined by each method of assessment weighted as follows:

- 6-7 weekly homework assignments  (55%)
- Midterm examination (15%, Wednesday, 10/23/2019)
- Final examination (30%, according to Stanford calendar: Wednesday, 12/11/2019 @ 3:30 PM, location TBD)

##  Policies

- Class Participation
    - This includes in-class participation and Canvas Discussion.
    - When homework involves simulations and data analysis, you will use R statistical computing software. Please post your R or R Markdown questions to Canvas Discussion.
    - Instructor or TA or other students in your class can answer your questions.
    - When asking questions about code, be specific (copy and paste the exact error, relevant code, and describe what you are attempting to do). We will not answer questions that are too similar to the problem sets or that would be better answered in office hours with a whiteboard. Needless to say, you should conduct yourself in a courteous and respectful manner on Canvas Discussion.

    
- Weekly homework assignments
    - Homework assignment will be assigned every Friday on Canvas Assignments.
    - Homework assignment will be due every Friday. 
        - Submit your completed homework assignment in PDF format, and submit a copy to your folder on canvas.
    - Use [R markdown](https://bookdown.org/yihui/rmarkdown/) in R Studio and render it to PDF
        - See the following [link](http://www.stat.cmu.edu/~cshalizi/rmarkdown/) for further outline of using [R markdown for reporting](http://www.stat.cmu.edu/~cshalizi/rmarkdown/rmarkdown.Rmd).
    - Each question in the homework assignment will be graded as follows: $scale \in \left\lbrace0,1,2,3,4\right\rbrace$ 
        - 4: submitted on time and more or less completely correct
        - 3: submitted on time and partially correct answer
        - 2: late submission and more or less completely correct
        - 1: late submission and partially correct answer
        - 0: late submission (any day after the due date for more than one homework assignment) or submitted with a completely incorrect answer
    - Each student can hand in only one homework late (within two days after the deadline). 
    - After attempting homework problems on an individual basis, students may discuss a homework assignment with their classmates. However, students must write up their own solutions individually and explicitly indicate from whom (if anyone) or resources students received help at the top of their homework solutions.
   
- Midterm examination 
  - In-class examination.
  - Midterm examination: \rc Wednesday, October 25, 2019 @ 1:30PM - 2.20 PM, Gates B3.\bc
  - Students **are not allowed** to take midterm examinations other than scheduled date (except the event of extraordinary circumstances that solely determined by the instructor)
    
- Final examination
  - In-class examination.
  - Following the Stanford [calendar](https://registrar.stanford.edu/autumn-quarter-exams): \rc Wednesday, December 11, 2019 @ 3:30PM-6:30 PM, location TBD.\bc
  - Students **are not allowed** to take final examinations earlier than the scheduled time (except the event of extraordinary circumstances that solely determined by the instructor).
  - Students **are not allowed** to take this course with another conflicting final examination. 
    
- Accessible Education
    - Students with Documented Disabilities: Students who may need an academic accommodation based on the impact of a disability must initiate the request with the Office of Accessible Education (OAE). Professional staff will evaluate the request with required documentation, recommend reasonable accommodations, and prepare an Accommodation Letter for faculty. Unless the student has a temporary disability, Accommodation letters are issued for the entire academic year. Students should contact the OAE as soon as possible since timely notice is needed to coordinate accommodations. The OAE is located at 563 Salvatierra Walk (phone: 723-1066, [URL:https://oae.stanford.edu/](https://oae.stanford.edu/).)
    - Provide me an accommodation letter on or before **09/27/2019**.
    
- Honor Code
    - Students are bound by the [Stanford Honor Code](https://communitystandards.stanford.edu/policies-and-guidance/honor-code). Violation of the honor code will result in a failing grade among other penalties.
    
- Stanford Center for Professional Development (SCDP)
  - Stats 191 is listed as one of the SCDP courses.
  - Lecture recordings are being made and might be shared with others at Stanford beyond those currently enrolled in the class.
  - SGDP policies on student privacy: __Video cameras located in the back of the room will capture the instructor presentations in this course. For your convenience, you can access these recordings by logging into the course Canvas site. These recordings might be reused in other Stanford courses, viewed by other Stanford students, faculty, or staff, or used for other education and research purposes. Note that while the cameras are positioned with the intention of recording only the instructor, occasionally a part of your image or voice might be incidentally captured. If you have questions, please contact a member of the teaching team.__

## Course Schedule

|Date|Week|Topic|Reading|Notes|
|--|--|--|--|--|
|**09/23/2019**|Week 1|--|--|--|
|**09/25/2019**|Week 1|--|--|--|
|**09/27/2019**|Week 1|--|--|--|
|**09/30/2019**|Week 2|--|--|--|
|**10/02/2019**|Week 2|--|--|--|
|**10/04/2019**|Week 2|--|--|--|
|**10/07/2019**|Week 3|--|--|--|
|**10/09/2019**|Week 3|--|--|--|
|**10/11/2019**|Week 3|--|--|--|
|**10/14/2019**|Week 4|--|--|--|
|**10/16/2019**|Week 4|--|--|--|
|**10/18/2019**|Week 4|--|--|--|
|**10/21/2019**|--|--|--|--|
|**10/23/2019**|--|--|--|\rc Midterm Examinations \bc |
|**10/25/2019**|--|--|--|--|
|**10/28/2019**|--|--|--|--|
|**10/30/2019**|--|--|--|--|
|**11/01/2019**|--|--|--|--|
|**11/04/2019**|--|--|--|--|
|**11/06/2019**|--|--|--|--|
|**11/08/2019**|--|--|--|--|
|**11/11/2019**|--|--|--|--|
|**11/13/2019**|--|--|--|--|
|**11/15/2019**|--|--|--|--|
|**11/18/2019**|--|--|--|--|
|**11/20/2019**|--|--|--|--|
|**11/22/2019**|--|--|--|--|
|**11/25/2019**|--|--|--|Thanksgiving Recess (no classes)|
|**11/27/2019**|--|--|--|Thanksgiving Recess (no classes)|
|**11/29/2019**|--|--|--|Thanksgiving Recess (no classes)|
|**12/02/2019**|--|--|--|--|
|**12/04/2019**|--|--|--|--|
|**12/06/2019**|--|--|--|--|
|**12/11/2019**||--|--|\rc End-Quarter examinations \bc |
```{r echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
schedule <- "
|Date     | Week  | Topic                      |Reading        |Notes     |
|---------|-------|----------------------------|---------------|----------|
| **04/01/2019**|Week 1 | Overview of current research in nonparametric and adequate initiation to R and R Markdown| [ASA Nonparametric statistics section news](https://community.amstat.org/nonparametricstatisticssection/news) gives an overview of some current research topics; install [R](https://www.r-project.org/); install [RStudio](https://www.rstudio.com/products/rstudio/); [TryR](http://tryr.codeschool.com/), [R Markdown webinar](https://www.rstudio.com/resources/videos/r-markdown-eight-ways/), [R Markdown](https://rmarkdown.rstudio.com/) provide adequate initiation to R and R Markdown.| |
|**04/03/2019**| Week 1| Logistics and Preliminaries | Syllabus, HWC: Chapter  1||
|**04/05/2019**| Week 1| The One-sample problem I (testing procedure) | HWC: Chapter 3.4-3.6, 3.8, 3.1-3.3, 3.7|**Homework 1 posted**|
|**04/08/2019**| Week 2| The One-sample problem II (estimator associated with the statistic, confidence interval, example) | HWC: Chapter 3.4-3.6, 3.8, 3.1-3.3, 3.7||
|**04/10/2019**| Week 2|Statistical functionals and Influence functions| Notes will be posted on W: Chapter 2, ET: Chapter 4, 5, 21.3||
|**04/12/2019**|Week 2| Jackknife and Bootstrap I  | HWC: Chapter 8.4 and notes will be posted based on W: Chapter 3, DH, ET: Chapter 6, 11|**Homework 2 Posted**, **Homework 1 Due**|
|**04/15/2019**| Week 3 | Bootstrap II |Notes will be posted based on ET: Chapter 23, Re:DH1994, Re:DH1995 ||
|**04/17/2019**| Week 3|Discrete data problems I  |HWC: Chapter 2||
|**04/19/2019**| Week 3| Discrete data problems II |HWC: Chapter 10|**Homework 3 Posted**, **Homework 2 Due**|
|**04/22/2019**| Week 4| Two-sample problem I | HWC: Chapter 4||
|**04/24/2019**| Week 4| Two-sample problem II| HWC: Chapter 5||
|**04/26/2019**| Week 4| Permutation Test I | |**Homework 4 Posted**, **Homework 3 Due**|
|**04/29/2019**| Week 5| Permutation Test II| | |
|**05/01/2019**| Week 5| Ranked-based linear regression| HWC: Chapter 9| 
|**05/03/2019**| Week 5| Smoothing I  |W: Chapter 4|**Midterm project proposal due** |
|**05/06/2019**| Week 6| Nonparametric regression I  |HWC: Chapter 9.7, 14, W: Chapter 5  | |
|**05/08/2019**| Week 6|Nonparametric regression II | HWC: Chapter 9.7, 14, W: Chapter 5 | |    
|**05/10/2019**| Week 6| Wavelets|HWC: Chapter 13, W: Chapter 9 | **Homework 5 Posted**, **Homework 4 Due**|
|**05/13/2019**| Week 7|ANOVA I| HWC: Chapter 6, 7|   |
|**05/15/2019**| Week 7|ANOVA II , multiple comparison|HWC: Chapter 6, 7 | | 
|**05/17/2019**| Week 7|Survival analysis I |HWC: Chapter 10 | **Homework 6 Posted**, **Homework 5 Due**| 
|**05/20/2019**| Week 8|Survival analysis II | HWC: Chapter 10| | 
|**05/22/2019**| Week 8|Ranked set sampling | HWC: Chapter 15| |
|**05/24/2019**| Week 8| Bayesian nonparametric I|HWC: Chapter 16, Li:W2016 |**Homework 7 Posted**, **Homework 6 Due** |
|**05/27/2019**| Week 9|**(Holiday, no classes)** | |Memorial Day |
|**05/29/2019**| Week 9|Bayesian nonparametric II | HWC: Chapter 16, Li:W2016 | | 
|**05/31/2019**| Week 9|Inference for data visualization | Re:BHLLSW2009, Re:D1983, Re:JWH2014|**Homework 7 Due** (no late submission allowed, End-Quarter Period starts) |
|**06/03/2019**| Week 10|Multivariate nonparamteric tests, Bootstrap III |ET: Chapter 12, 14| |
|**06/05/2019**| Week 10|Wrap-up  | | **Final project due**|
"
cat(schedule)
```



##  Important Dates

|Date|Day|Description|
|--|--|--|
|10/11/2019| Friday, 5:00 p.m. |Last day to add or drop a class|
|10/23/2019| Wednesday, 1:30-2:20 p.m. |Midterm examination|
|11/04/2019 |Monday, 5:00 p.m.   | Term withdrawal deadline with a partial refund|
|11/25/2019- 11/29/2019 |Monday - Friday  | Thanksgiving Recess (no classes)  |
|12/02/2019 - 12/08/2019 | Monday - Sunday | End-Quarter Period |
|12/06/2019 |Friday| Last day of classes                     |
|12/11/2019 |Wednesday| Final Examinations @ 3:30 p.m. - 6.30 p.m.|    
|12/17/2019 |Tuesday, 11.59 p.m.| Grades due|



_Disclaimer: This syllabus provides a general plan; deviations may be necessary._

